---
title: "**DATA303 Interim Report - Group 5**"
author:
- 'Michael Fry 300570669'
- 'Fletcher Smith '
- 'Matthew Smaill '
date: "`r format(Sys.time(), '%A, %d %B')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
toc: yes
bibliography: Interim_Report.bib
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---

\newpage

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      echo = FALSE, 
                      fig.width=8, 
                      fig.height=4) 

# To install all third party packages at once
#install.packages(c('dplyr','ggplot2','ggcorrplot','tidyr', 'pander', 'flextable')))

library(ggcorrplot)
library(dplyr)
library(ggplot2)
library(tidyr)
library(pander)
library(flextable)
options(scipen = 999)

# Ensure that the dataset is stored in this directory, it is too large to commit to Github so needs to be downloaded and stored locally in this directory.
crash_data = read.csv('../../Data/crash_data.csv')

```

# Background and Data

## Background

The New Zealand Crash Analysis System (CAS) dataset is a comprehensive compilation of traffic crash information recorded by the Waka Kotahi, the New Zealand Transport Agency. The CAS dataset constitutes a valuable resource for gaining insights into the factors contributing to traffic crashes across New Zealand's diverse roadways and public-access areas. This report provides an overview of the CAS dataset, its significance, contents, and inherent characteristics.

## Dataset (source and coverage)

The CAS data originates from the Waka Kotahi Crash Analysis System, which serves as a repository for all reported traffic crashes involving motor vehicles in New Zealand. This system is only fueled by information provided by the New Zealand Police. The scope of CAS encompasses crashes that occur on any road segment or area within the country where the public has legal access with a motor vehicle. This extensive coverage ensures that the dataset represents a wide array of scenarios, road types, and conditions.


## Importance

The CAS dataset is of considerable interest due to its potential to address critical questions surrounding road safety and accident prevention. One of the central questions that this 
dataset can help answer is: "What statistical techniques can we use to find the relational effect that key variables have on major automotive crashes?" By analyzing the dataset, we  can 
identify patterns, correlations, and trends that shed light on the factors contributing to major vehicle crashes. 

## Types of Data 

The CAS dataset incorporates various types of data, each contributing to a holistic understanding of traffic crashes. This dataset comprises 12 logical variables, 2 date variables, 15 
categorical variables, and 41 numeric variables. The inclusion of diverse data types allows for a multi-faceted analysis that captures both quantitative and qualitative aspects of crash 
incidents.

## Completeness 

It's important to note that the CAS dataset, while comprehensive, does contain missing values. Out of all the columns in the dataset, only X, Y, ObjectID, and crashYear are entirely devoid 
of missing values. However, various other variables exhibit significant instances of missing data. For instance, variables such as Bridge, debris, fence, vehicle, and waterRiver each have 
488,831 missing values. This variation in missing data across variables underscores the complexity of real-world data collection and emphasizes the need for careful consideration when 
conducting analyses or drawing conclusions.

In conclusion, the New Zealand Crash Analysis System (CAS) dataset serves as a valuable resource for investigating the dynamics of traffic crashes in New Zealand. Its extensive coverage, 
diverse data types, and potential to answer crucial questions make it an essential tool for researchers, policymakers, and analysts aiming to enhance road safety and prevent major 
automotive crashes. However, the presence of missing data underscores the importance of thorough data preprocessing and analysis techniques to ensure accurate and meaningful insights.



\newpage

# Ethics and Privacy 

Privacy and ethical considerations are paramount for data scientists when dealing with, and modelling data. Especially sensitive data. This section of the report details the privacy and ethical concerns that must be addressed when dealing with the crash (CAS) dataset to answer the question: Can we use statistical analysis techniques to find the relational effects that key factors have on leading to a major crash using the CAS dataset? It also discusses what steps could be taken to ensure the project data and results are secure. 
Privacy considerations refers to the protection of individuals' personal information and sensitive data. It involves ensuring that data is collected, stored, processed, and shared in a way that respects individuals' rights and maintains confidentiality. Ethical considerations on the other hand, is all about how the data is being used, especially when it comes to potential biases, fairness, and social implications.
It is important to note; the CAS dataset is publicly available and provided by a government entity. Therefore, the following discussion includes BOTH methods that the NZTA has done to try and privatise the data, AND also the privacy and ethical considerations that still need to be taken  into account within this project. 


## Privacy 

Privacy considerations are of utmost importance when dealing with any dataset, especially one as sensitive as the Crash Analysis System (CAS). Personal identifiers, including the gender and age of the driver, and license plate numbers, have been deliberately removed from the dataset by the NZTA. These removals are critical steps taken to protect individual privacy. However, even though these direct identifiers have been removed, this alone may not suffice to guarantee privacy.
The dataset still contains information about location and the financial year in which the crash occurred. While not as overt as personal identifiers, this information can still be used, in combination with other factors, to potentially identify individuals involved in crashes. Location data, in particular, can provide insights into the vicinity of the crash, which may be sensitive information.
It is import for us data scientists to consider this, as it could impact peoples lives, as our analysis hasn’t come with the direct permission with the participants for the use of their data (indirectly has been through the NZTA).

## Ethics 

Ethical considerations encompass a broad spectrum of concerns that go beyond mere data protection, extending to issues of fairness, bias mitigation, and the broader social implications of data analysis. Given that the CAS dataset is publicly accessible and provided by a government entity, the New Zealand Transport Agency (NZTA), ethical scrutiny must encompass an evaluation of the effectiveness of the NZTA's data privatization efforts. It is imperative to determine whether these efforts are robust enough to adequately safeguard individuals' privacy and whether the dataset aligns with ethical data sharing standards.
Furthermore, despite the public availability of the CAS dataset, ethical concerns emerge regarding the consent of individuals whose data is included. Ensuring that the data used in research has been obtained and shared in accordance with informed consent principles is a cornerstone of ethical data analysis, particularly when dealing with sensitive crash data. This might be the case for individuals when they consent for the use of this data by NZTA, however, we have not asked for consent directly by any crash participants. Therefore, we need to do our crash analysis with respect of the people whose data we are using. 

Analyzing crash data carries significant societal implications, potentially influencing policy decisions and public perceptions. Ethical considerations require data scientists to approach our analysis with a commitment to minimizing harm and promoting societal well-being. We should carefully consider how their findings may impact communities and advocate for responsible data use, recognizing the sensitivity of the information involved.



## Steps to Ensure Privacy and Ethical Compliance:

1.	Data Anonymization: The first step involves a rigorous review of the dataset to eliminate or anonymize any remaining personally identifiable information. This process aims to further protect individuals' privacy by removing the possibility of identification.

2.	Adding Noise: To bolster privacy protection, consider adding noise to sensitive attributes like crash dates. This technique obscures specific details about individuals while still allowing meaningful analysis, thereby striking a balance between utility and privacy. If the location variables, where to be used in our analysis, then we could move the area of crash within a 10km radius, which wouldn’t effect the overall results of our findings.  

3.	Ethical Review: Conduct a comprehensive ethical review of the research project. This review should identify potential biases and ethical dilemmas, requiring consultation with experts in ethics and data privacy to ensure responsible research.

4.	Fairness Assessment: Scrutinize the dataset for fairness issues, such as disparities in crash reporting or data collection. Mitigate any identified biases during data analysis and reporting to ensure fairness in the research.


\newpage

# Exploratory Data Analytics


```{r}

# Summarize the data
fatalities_by_vehicle <- crash_data %>%
  summarise(
    Bicycle = sum(bicycle * fatalCount, na.rm = TRUE),
    Bus = sum(bus * fatalCount, na.rm = TRUE),
    Car_StationWagon = sum(carStationWagon * fatalCount, na.rm = TRUE),
    Moped = sum(moped * fatalCount, na.rm = TRUE),
    Motorcycle = sum(motorcycle * fatalCount, na.rm = TRUE),
    SUV = sum(suv * fatalCount, na.rm = TRUE),
    Taxi = sum(taxi * fatalCount, na.rm = TRUE),
    Truck = sum(truck * fatalCount, na.rm = TRUE),
    Unknown_Vehicle = sum(unknownVehicleType * fatalCount, na.rm = TRUE),
    Van_Utility = sum(vanOrUtility * fatalCount, na.rm = TRUE)
  )

# Convert to a longer format
longer <- fatalities_by_vehicle %>% 
  pivot_longer(
    everything(),
    names_to = 'Type',
    values_to = 'Fatalities'
  )

## Group by Region 
total_crashes_by_region <- crash_data %>%
  group_by(region) %>%
  summarise(total_crashes = n())

# Step 2: Calculate the proportion of each type of crash within each region
crash_severity_by_region <- crash_data %>%
  group_by(crashSeverity, region) %>%
  summarise(n = n()) %>%
  left_join(total_crashes_by_region, by = "region") %>%
  mutate(proportion = n / total_crashes)

crash_severity_by_region <- crash_severity_by_region %>% 
    filter(!is.na(region) & region != "")

crash_severity_by_region$crashSeverity <- factor(crash_severity_by_region$crashSeverity, 
      levels = c("Non-Injury Crash", "Minor Crash", "Serious Crash", "Fatal Crash"))

crash_data$crashSeverity <- factor(crash_data$crashSeverity,  levels = c("Non-Injury Crash", "Minor Crash", "Serious Crash", "Fatal Crash"))

severity_colors <- c("Non-Injury Crash" = "#8DB3AD", "Minor Crash" = "#F5D7B1", "Serious Crash" = "#F05161", "Fatal Crash" = '#262626')
```


## Summary Tables 

```{r}
summary_table <- data.frame(
  Aspect = c("Number of Instances", 
             "Number of Features", 
             "Number of Missing Values", 
             "Number of Numeric Features", 
             "Number of Categorical Features"),
  Value = c(nrow(crash_data), 
            ncol(crash_data), 
            sum(is.na(crash_data)), 
            length(select_if(crash_data, is.numeric)),
            (ncol(crash_data) - length(select_if(crash_data, is.numeric))))
)

# Displaying the summary table
pander(summary_table)
```
This table shows the count of Instances, Features, Missing values, Numerical and Categorical features within the CAS Dataset. Please note, these values are different than those reported in the Background section, as it is likely that some variables will need ot be re-encoded at the data cleaning stage of the analysis.

```{r}
# Create a table showing the count and percentage of missing values for each variable
missing_value_table <- data.frame(
  Count = sapply(crash_data, function(x) sum(is.na(x))),
  stringsAsFactors = FALSE
) %>%
  mutate(Percent_Missing = (Count / nrow(crash_data)) * 100) %>%
  filter(Percent_Missing > 10) %>%
  arrange(desc(Percent_Missing))



pander(missing_value_table)
```

This table shows the count and percentage of missing values for each variable in the dataset with more than 10% of the data missing. The table provides insights into the completeness of the dataset. Columns with higher percentages of missing values may require further investigation or data imputation strategies.

\newpage



## Plots 

### Speed Limit Distribution



```{r,fig.cap = 'Histogram for distribution of advisory speeds in the dataset.  The speed limit distribution shows the distribution shows the number of crashes at different speed limits. (Note, this is not the speed of the crash, just the speed limit at the location of the crash). The most common speed limit that crashes occur at is 50km/h, followed then by 100km/h. This isnt very surprising. Firstly, the default urban speed limit 50km/h, with a high volume of traffic driving under this speed limit across the country each day. Further, this would also included other types of crashes, such as bikes and busses. This then makes sense that it experiences the most crashes. Majority highways and open roads in New Zealand have a speed limit of 100, and if cars where going in excess of 100km/h then these crashes are also included within the 100km/h range.'}
ggplot(crash_data, aes(x = speedLimit)) + 
  geom_histogram(binwidth = 5) +
  theme_minimal() +
  labs(x = "Speed Limit (Km/h)", y = "Count", title = "Speed Limit Distribution")+ 
  scale_x_continuous(breaks = seq(0, 100, by = 10))

```
From the graph above;


### Advisory Speed Distribution

```{r,fig.cap ='The histogram for speed limits shows a similar trend to the advisory speed, where most of the crashes occur around the 50-60 km/h range.'}
ggplot(crash_data, aes(x = advisorySpeed)) + 
  geom_histogram(binwidth = 5) +
  theme_minimal() +
  labs(x = "Advisory Speed (Km/h)", y = "Count", title = "Advisory Speed Distribution") + 
  scale_x_continuous(breaks = seq(0, 100, by = 10))
```





### Boxplot of advisorySpeed vs crashSeverity

```{r ,fig.cap = 'The boxplot between advisory speed and crash severity shows that the advisory speed tends to increase slightly with the severity of the crash. The notches in the fatal crash boxplot suggests that there is a statistically significant difference between the medians of Non-injury, minor and serious crashes with fatal crashes. This plot may imply that higher advisory speeds are associated with more severe crashes, although correlation does not imply causation.'}
ggplot(crash_data, aes(x = crashSeverity, y = advisorySpeed, fill = crashSeverity)) + 
  geom_boxplot(notch=TRUE) +
  ggtitle("Advisory Speed vs Crash Severity") +
  xlab("Crash Severity") + 
  ylab("Advisory Speed (Km/h)") + 
  scale_fill_manual(values = severity_colors)+ 
  scale_y_continuous(breaks = seq(0, 110, by = 10))
  
```



### Boxplot of speedLimit vs crashSeverity

```{r, fig.cap ='The boxplot between Speed Limit and crash severity shows some of the same relationship as the advisory speed graph above.'}
ggplot(crash_data, aes(x = crashSeverity, y = speedLimit, fill = crashSeverity)) + 
  geom_boxplot(notch=TRUE) +
  ggtitle("Speed limit vs Crash Severity") +
  xlab("Crash Severity") + 
  ylab("Speed Limit (Km/h)") + 
  scale_fill_manual(values = severity_colors)+ 
  scale_y_continuous(breaks = seq(0, 110, by = 10))

```


### Number of lanes 

```{r, fig.cap ='The bar plot shows that most crashes occur on roads with 1 or 2 lanes.This makes seance since the majority of crashes happen in 50km/h zones witch are normally in towns and cities.' }
ggplot(crash_data, aes(x = NumberOfLanes)) + 
  geom_bar() +
  ggtitle("Distribution of Number of Lanes") +
  xlab("Number of Lanes") + 
  ylab("Count")+ 
  scale_x_continuous(breaks = seq(0, 10, by = 1))

```



### Barplot of Fatalities by Vehicle Type

```{r, fig.cap ='The bar plot shows the number of fatalities by vehicle type. Its clear that cars are the most common vehicle to be involved in an fatal crash compared to all other vehicle types. this is expected sine light passenger vehicles accounted for 76.4% of the vehicles on the road. One interesting observation is the fact that there are more trucks involved in fatal crashes than motorcycles when there are more motorcycles on the road. This could be due to trucks being driven more often and they can cause a lot more damage due to their size/ weight compared to motorcycles.'}

# Create the barplot
ggplot(longer, aes(x = reorder(Type, -Fatalities), y = Fatalities)) + 
  geom_bar(stat = 'identity') +
  ggtitle("Fatalities by Vehicle Type") +
  xlab("Vehicle Type") + 
  ylab("Number of Fatalities")+ 
  theme(axis.text.x = element_text(angle = 90))
```




### Fatalities by year 

```{r, fig.cap ='The line plot shows a clear trend in the number of fatalities. The period between 2000 and 2013 showes a clear downward trend of the number of fatalities. The period between 2013 and 2022 (Excluding 2023 as there is not data for the whole year) there is a clear upward trend in number of fatalities. This could be due to numerious things, population growth, increased reporting or other factors.'}
fatalities_by_year <- crash_data %>%
  group_by(crashYear) %>%
  summarise(Total_Fatalities = sum(fatalCount, na.rm = TRUE))

ggplot(fatalities_by_year,aes(x = crashYear, y = Total_Fatalities)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Fatalities by Year", y = "Number of Fatalities", x='Year of Crash')+ 
  scale_x_continuous(breaks = seq(2000, 2023, by = 2))+ 
  scale_y_continuous(breaks = seq(0, 500, by = 50))

```



### Fatalities by weather A 

```{r, fig.cap = 'The bar plot shows the number of fatalities categorized by primary weather conditions. Its clear that fatal car crashes are most likely to happen when the weather is fine. This is somewhat surprising considering it rains 111 days per year on average in New Zealand. However, it could be due to people driving more taking risks when the weather is better.'}
fatalities_by_weather <- crash_data %>%
  group_by(weatherA) %>%
  summarise(Total_Fatalities = sum(fatalCount, na.rm = TRUE))

ggplot(fatalities_by_weather, aes(x = reorder(weatherA, -Total_Fatalities), y = Total_Fatalities)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Fatalities by Weather Condition (A)", y = "Number of Fatalities", x='Weather Condition A') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


### Fatalities by weather B 

```{r, fig.cap ='Similar to weatherA, this bar plot shows fatalities by secondary weather conditions.'}
fatalities_by_weatherB <- crash_data %>%
  group_by(weatherB) %>%
  summarise(Total_Fatalities = sum(fatalCount, na.rm = TRUE))%>%
  filter(weatherB != 'Null')

ggplot(fatalities_by_weatherB, aes(x = reorder(weatherB, -Total_Fatalities), y = Total_Fatalities)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Fatalities by Weather Condition (B)", y = "Number of Fatalities", x='Weather Condition B' ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```







### Crash Severity

```{r, fig.cap = 'The bar plot gives an overview of the crash severity distribution in the dataset. Most crashes are non-injury or minor, but there is a significant number of serious and fatal crashes as well. This is probably due to the fact that most crashes occur when the speed limit is 50km/h or below and this is a relatively low speed. Another reason for the high number of non-injury crashes could be due to the high number of "fender benders" that are reported.'}

ggplot(crash_data, aes(x = crashSeverity, fill = crashSeverity)) +
  geom_bar() +
  theme_minimal() +
  labs(x = "Crash Severity", y = "Count", title = "Crash Severity Counts") + 
  scale_fill_manual(values = severity_colors)

```

### Crash Severity by Region 

```{r,fig.cap ='The bar plot shows the regional distribution of proportion of crashes at each crash severity. This can give insights into which regions have a higher proportion of more severe crashes. The proportions between non-injury, minor, serious  and fatal crashes across all regions are relatively similar, with slight variations between them. Only a small proportion of crashes are fatal (around 2%, +-0.05%), compared to Non-injury crashes (around 60%).However, as the crash severity gets worse, there is slightly more variation in the proportion of each crash type between regions.  Even though some regions have slightly more server crashes than other regions (such as West coast and Northland); It is clear that for our analysis, there is no one particular region that has an alarmingly higher proportion of serious/fatal crashes than other regions.'}
# Create the first plot
ggplot(crash_severity_by_region, aes(x = region, y = proportion, fill = crashSeverity)) +
  geom_bar(stat = 'identity') +
  facet_wrap(~crashSeverity, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0)) +
  labs(title = 'Proportion of Crash Severity by Region',
       x = 'Region',
       y = 'Proportion of Crashes') +
  scale_fill_manual(values = severity_colors)  # Apply custom colors
```



### Crash Severity by Region - Stacked

```{r, fig.cap ='The stacked bar plot shows the same regional distribution of the proportion of crashes at each severity as above. The stacked nature of the bars allows for a different view of the data, providing an easier way to compare regions with one plot.'}
# Create the second plot
ggplot(crash_severity_by_region, aes(x = region, y = proportion, fill = crashSeverity)) +
  geom_bar(stat = 'identity') +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(
    title = "Proportion of Crash Severity by Region",
    x = "Region",
    y = "Proportion"
  ) +
  scale_fill_manual(values = severity_colors) 

```



### Heatmap of Fatal Crashes by year and region.

```{r, fig.cap ='The heatmap shows the distribution of total fatal crashes by year and region. Darker colors represent higher numbers of fatal crashes. The waikato region has had continual high levels of fatal crashs. This peaked in the early 2000. Based on the Waikato Regional Road Safety Report in From the graph above; The Waikato region has had continual high levels of fatal crashs. This peaked in the early 2000. Based on the Waikato Regional Road Safety Report 2000, "The four main contributing factors in all rural crashes in the Region were poor observation, excess speed, road factors and alcohol". This is useful for our anaysis, as it shows four potential causes of a fatal crash. The Nelson Region and the Marlborough on the other hand, had very low levels of fatal crashes. It is interesting to see, that the level of fatal crashes within each region does stay relatively the same overtime, with only some small changes within each region overtime. , "The four main contributing factors in all rural crashes in the Region were poor observation, excess speed, road factors and alcohol". This is useful for our anaysis, as it shows four potential causes of a fatal crash. The Nelson Region and the Marlborough on the other hand, had very low levels of fatal crashes. It is interesting to see, that the level of fatal crashes within each region does stay relatively the same overtime, with only some small changes within each region overtime.'}

fatal_summary <- crash_data %>%
  filter(crashSeverity == "Fatal Crash") %>%
  group_by(crashYear, region) %>%
  summarise(total_fatal = sum(fatalCount)) %>%
  ungroup()

ggplot(fatal_summary, aes(x=crashYear, y=region)) +
  geom_tile(aes(fill=total_fatal)) +
  scale_fill_gradient(low="white", high="red") +
  ggtitle("Heatmap of Total Fatal Crashes by Year and Region") +
  xlab("Year") +
  ylab("Region")


```


### Things to note going forward

There are a significant number of variables with missing data. Each will need to be handled differently. Significant pre-processing will be needed before the fitting of any models. 

It is likely, due to the significant number of features present in the dataset, that significant feature selection will be required to reduce the dimensionality of the dataset to a usable level. 



\newpage


# Individual Contributions

## Michael Fry 

- Created and formatted the interm report document 
  - Knittr options, layout, formatting etc. 
- Exploratory Data Analytics Section
  - Summary
  - Grouping etc
  - Plots 
  - Captions

## Fletcher Smith 
- Background and Data
- EDA plot descriptions

## Matthew Smaill
- Data and privacy 
- EDA plot descriptions

# Link to github
https://github.com/MF783/DATA301-Group-5
\newpage

# References 


